\input{../templates/slides_template}
\input{../templates/php_info}

% meta-information
\newcommand{\topic}{
	% TODO fill in the actual topic
	PHP-Einführung - Lesson 2 - Control Flow
}

\title{\topic}
\date{\today}

% the actual document
\begin{document}

\maketitle

\begin{frame}{Content of this lesson}

	\setbeamertemplate{section in toc}[sections numbered]
	\tableofcontents

\end{frame}

\section{Data Types}

\begin{frame}[fragile]{Data types}

	Though PHP does not require explicit type declaration, it evaluates the type of a variable based on it's content: \pause
	\begin{lstlisting}
<?php
$name = "Heinz"; //String
$answer = 42; //Integer
$pi = 3.14159265359; //Float/Double
$isTrumpAGoodPresident = false; //Boolean
$goodFirstNames = ["Jakob","Markus"]; //Array
$currentDate = new DateTime(); //Object
$whyIsThisNull = null; //NULL
$file = fopen("info.txt"); //Resource
	\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{Type casting}

	Type casting works much as it does in other high-level languages like Java or C++\pause
	\begin{lstlisting}
<?php
$answer = 42; //Integer
$isThisTheRealAnswer = (boolean) $answer; 
var_dump($isThisTheRealAnswer); //prints true
	\end{lstlisting}
	\pause
	
	\textbf{HINT:} The \emph{var\_{}dump(\$variable)} function is used to print out the value(s) of a variable. It is very useful for debugging ;)

\end{frame}

\begin{frame}[fragile]{Type juggling}

	It will become a bit weird now. PHP automatically \emph{converts} variable types, \emph{based on the operations performed on them}.\pause
	\begin{lstlisting}
<?php
$foo = "0";  // $foo is currently a string, respresenting the char "0"
$foo = $foo + 2;   // $foo is now an integer with the value 2
$foo = $foo + 1.3;  // $foo is now a float (3.3)
$foo = 5 + "10 Little Piggies"; // $foo is an integer (15)
?>
	\end{lstlisting}
\end{frame}

\begin{frame}[fragile]{Constants}

There are also constants in PHP! These are declared by using the \emph{define()} function and are globally accessible: \pause

\begin{lstlisting}
<?php
define("GREETING", "Hello students!");
echo GREETING;
\end{lstlisting}
\pause

Constants can also be case-insensitive: \pause
\begin{lstlisting}
<?php
define("goodbye", "See you again students!", true);
echo GooDbYE;
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{String parsing}

There are some minor differences between double and single-quote strings: \pause

\begin{lstlisting}

<?php
$type = "cookie";

//This will evaluate the $type variable and print out the "correct" string
echo "He drank some coffee with $type syrup";
//While this echo will not, because variables are not evaluated in single-quote strings
echo 'He drank some coffee with $type syrup';

//Invalid, because the variables name is "type" and not types.. What now?
echo "But the syrup was not made out of $types";

//You can specify the end of the variable name by enclosing it in braces:
echo "But the syrup was not made out of ${type}s";
\end{lstlisting}

\end{frame}

\section{Operators}

\begin{frame}[fragile]{Arithmetic Operators}

While PHP has a lot of operators, let's cover the arithmetic ones first \pause

\begin{lstlisting}
<?php
$a = 14;
$b = 7;

//Calculation
$addition = $a + $b;
$subtraction = $a - $b;
$multiplication = $a * $b;
$division = $a / $b;
$modulus = $a % $b;
$exponentiation = $a ** $b; //Since 5.6

//Concatenation
$first = "Alex";
$last = "Lichter";
$fullname = $first . " " . $last;
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{Assignment Operators}

PHP also delivers some shortcuts to assign variables (similar to other languages):\pause

\begin{lstlisting}
<?php
$result = 14;
$otherNumber = 7;

//Calculation shorthands
$result += $otherNumber; //Same as $result = $result + $otherNumber
//Works with all arithmetic operators the same way 

$otherNumber++; //short for $otherNumber += 1
$otherNumber--; //short for $otherNumber -= 1

//Concatenation
$fullname = "Alex";
$fullname .= " Lichter";
//Same pattern as above
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{Comparison Operators}

For the control flow of a program, we need to compare variables:\pause
\begin{lstlisting}
<?php
$foo = "0"; $bar = 0;

//Equal
var_dump($foo == $bar); //true -> Equal value (though it's not the same type)
var_dump($foo == false); //true -> Same as above
//Not equal
var_dump($foo != $bar); //false

//Identical
var_dump($foo === $bar); //false -> Equal value, but not the same type
//Not identical
var_dump($foo !== $bar); //true

var_dump(1 >= 1); //true
var_dump(2 < 3); //true
//The typical greater/less than operators are available 
//>=, >, <, <=
\end{lstlisting}

\end{frame}

\begin{frame}[fragile]{Logical Operators}

We can combine booleans by using logical operators\pause
\begin{lstlisting}
<?php
$a = false; $b = true;

//OR (non-exclusive)
var_dump($a || $b); //true -> at least one statement is true
var_dump($a or $b);
var_dump($a || $a); //false

//XOR
var_dump($b XOR $b); //false -> exclusive or (At least one, but not both)

//AND
var_dump($a && $b); //false -> only true when both values are true
var_dump($a && true); //true

//NOT
var_dump(!$a); //true
\end{lstlisting}

\end{frame}

\section{If/Else}
\begin{frame}[fragile]{Control Flow - ITE \#{}1}

Before you get your first task today, let's talk about conditions and ITE.\pause

ITE means \textbf{I}f \textbf{T}hen \textbf{E}lse and is necessary for structuring your program. A little example here: \pause
\begin{lstlisting}
<?php
$hour = date("H"); //Returns the current hour

if($hour < 18){
	echo "Good day ladies and gentlemen!";
}
\end{lstlisting} 
\pause

What does that little script do? \pause

Exactly! The script evaluates if the current hour is less than 18, and prints out the sentence if the condition is met!

\end{frame}

\begin{frame}[fragile]{Control Flow - ITE \#{}2}

But we should also print something out when it's evening, right?
How would we do that? \pause
\begin{lstlisting}
<?php
$hour = date("H"); //Returns the current hour

if($hour < 18){
	echo "Good day ladies and gentlemen!";
} else {
    echo "Good evening ladies and gentlement!";
}
\end{lstlisting} 
\pause

Yup! This is how if then else works.

\end{frame}

\begin{frame}[fragile]{Control Flow - ITE \#{}3}

And.. what should we do when we want to print out something when the hour is equal to 20? \pause
\begin{lstlisting}
<?php
$hour = date("H"); //Returns the current hour

if($hour < 18){
	echo "Good day ladies and gentlemen!";
} else if($hour == 20) {
    echo "Primetime is starting soon!";
} else {
    echo "Good evening ladies and gentlement!";
}
\end{lstlisting} 
\pause

Now, the script will grab the current hours and go through all conditions. When one of them is true, the code inside it will be executed and no other conditions will be verified.
\end{frame}

\begin{frame}[fragile]{Control Flow - ITE \#{}4 - Your task}

Alright! Now you get your first little exercise to accomplish. \pause

The easter bunny went through a lot of gardens to hide \emph{precious eggs} a few days ago! There were so many gardens that he used a script that told him \emph{how many eggs} he should hide in each garden. Glad that \textbf{you} wrote him a great PHP script!

For the families "Müller", "Härtel" and "Lehner", there were \emph{12} eggs to hide. 

For the "Baumann", "Noack" and "Bodirsky" families, there were even \emph{17} eggs to hide.

But for the families "Fischer", "Neumann" and "Haustein" only \emph{5} eggs. All other families had the same number of eggs: \emph{7}

(Because we hadn't talked about how to use input forms yet, assume that you use a variable as input) 
\end{frame}

\section{Switch/Case}

\begin{frame}[fragile]{Control Flow - Switch/Case}

Because we do not want to end up with code like this:
\begin{lstlisting}
<?php
$number = 3; //random number here

if($number == 1){
	//Do sth.
} else if($number == 2) {
    //Do sth. else 
} else if($number == 3) {
    //Do sth. else again
} else if($number == 4) {
    .....
} else {
    //Something else again...
}
\end{lstlisting} 
\pause

we will use something called "Switch/Case". It's basically like ITE, but focused on one variable (the "switch")

\end{frame}

\begin{frame}[fragile]{Control Flow - Switch/Case}

Let's refactor our code from above
\begin{lstlisting}
<?php
$randomNumber = 3; //random number here
switch($randomNumber){
	case 1: {
		//Do sth. here
		break;
	}
	case 2:
	case 3: {
		//Do the same thing in case the var is 2 or 4
		break;
	}
	case 5: {
		//Do sth. else again
		break;
	}
	default: {
		//When no case matches
		break;
	}
}
\end{lstlisting} 
\pause

Looks much cleaner, doesn't it?

\end{frame}

\section{Loops}

\begin{frame}[fragile]{Loops}

Imagine that we want to output the same string 5 times.. \pause

\begin{lstlisting}
<?php
$output = "LOL";

//Echo 5x
echo $output;
echo $output;
echo $output;
echo $output;
echo $output;
\end{lstlisting} 
\pause

But.. that should be easier, shouldn't it?

\end{frame}


\begin{frame}[fragile]{Loops - While loop}

I want to introduce the first loop type: The \textbf{while loop}.
It will execute the code in the body while the condition argument is true  \pause

\begin{lstlisting}
<?php
$ouput = "LOL";

$count = 0;
while($count <= 5){
	echo $output;
	$count++; //Short form for $count +=1
}
\end{lstlisting} 

\end{frame}

\begin{frame}[fragile]{Loops - For loop}

But there is another type of loop that fits a bit more in this case: The \textbf{for loop}.
It is a bit more complicated than the while loop, but intuitive anyway\pause

\begin{lstlisting}
<?php
$ouput = "LOL";

for($count = 0; $count <= 5; $count++){
	echo $output;
}
\end{lstlisting} \pause

Can you see the similarities to the while loop?

It's basically the same, but with the variable initialization in the loop itself.
\end{frame}

\begin{frame}[fragile]{Loops - Break and continue}

Of course there are ways to stop a loop\pause

\begin{lstlisting}
<?php
for($count = 0; $count <= 5; $count++){
	echo "Hey";
	if($count == 3){
		break; //Stops the loop
	}
}
\end{lstlisting} \pause

Or to ignore some loop steps \pause

\begin{lstlisting}
<?php

for($count = 0; $count <= 5; $count++){
	if($count == 3){
		continue; //Ignores all code below
	}
	echo "Hey!";
}
\end{lstlisting} \pause

\end{frame}

\begin{frame}[fragile]{Loops - The next task}

Another task for you:

The easter bunny forgot to visit Egypt! And it came even worse. He don't know how a pyramid looks like! Please, write him a script that outputs a well-aligned pyramid out of a character of your choice! \pause

\textbf{HINT:} Wrap your pyramid code in "\textless{}pre\textgreater{}" HTML tags, so the spaces will be rendered correctly!

\end{frame}

\section{Arrays}

\begin{frame}[fragile]{Arrays}

Our last topic for today is called "Arrays". But what are arrays? \pause
Arrays are special variables that hold multiple values at once!

Imagine you have a list of cars.
\begin{lstlisting}
<?php
$car1 = "VW";
$car2 = "BMW";
$car3 = "Porsche";
//and so on..
\end{lstlisting} \pause

That doesn't look good! How could you loop through the car list now.. And imagine all the car types in one list... \pause

The solution is an \textbf{array}

\end{frame}


\begin{frame}[fragile]{Arrays}

\begin{lstlisting}
<?php
$cars = array("VW", "BMW", "Porsche");
//Or the shorthand:
$cars = ["VW", "BMW", "Porsche"];
\end{lstlisting} \pause
Great! Now you need to access the variables stored in the array somehow.. \pause

\begin{lstlisting}
<?php
$cars = ["VW", "BMW", "Porsche"];
echo "My car is a " . $cars[2]; //Outputs "My car is a Porsche"
\end{lstlisting} \pause

You access the variables by using its index (starting with 0).

\end{frame}

\begin{frame}[fragile]{Arrays}

There is also a way to count all elements of an array: \pause

\begin{lstlisting}
<?php
$cars = ["VW", "BMW", "Porsche"];
echo count($cars);
\end{lstlisting} \pause

To loop through an array, you could use the count and a for loop, as you have learned: \pause

\begin{lstlisting}
<?php
$cars = ["VW", "BMW", "Porsche"];
for($x = 0; $x < count($cars); $x++) {
    echo $cars[$x]. "\n";
}
\end{lstlisting} \pause

\end{frame}

\begin{frame}[fragile]{Arrays - Foreach loop}

But as you might already guessed, there is a shorter way for it! The \textbf{foreach loop} \pause

\begin{lstlisting}
<?php
$cars = ["VW", "BMW", "Porsche"];
foreach($cars as $carName) {
    echo "$carName\n";
}
\end{lstlisting} \pause

The loop iterates over all array elements! You can even do something with the indexes: \pause

\begin{lstlisting}
<?php
$cars = ["VW", "BMW", "Porsche"];
foreach($cars as $index => $carName) {
    echo "Position: $index | Car name: $carName\n";
}
\end{lstlisting} \pause

\end{frame}

\begin{frame}[fragile]{Associative Arrays}

Arrays don't need to have numeric indexes though. Those who have numeric indexes are called \emph{Indexed Arrays}, but there are also \emph{Associative Arrays}\pause

As the name already hints, they associate any index with an assigned value:
\begin{lstlisting}
<?php
$ages = []; //empty array
$ages["Alex"] = 20;
$ages["Peter"] = 9;
$ages["Inge"] = 86;
foreach($ages as $name => $age) {
    echo "$name ist $age Jahre alt\n";
}
\end{lstlisting} \pause

\end{frame}

\begin{frame}[fragile]{Arrays - Your final task}

As your final task for today, you should partly refactor the programs you already wrote before! \pause

Make a list (or many, as you wish) with all families that received a special number of eggs in Task \#{}1.\pause There are 48 more families that the easter bunny visited while carrying the same bag of eggs. \pause

Calculate how many eggs the easter bunny needs to put in his bag before visiting all families (including the "special" ones). \pause

Also, use knowledge from all sections from today, if possible or needed. \pause

\emph{BONUS:} Bring up some statistics about the egg distribution. ;)

\textbf{Good luck!}

\end{frame}


\end{document}
